---
title: "Use of duckplyr in other packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{developers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || rlang::is_installed(c("conflicted", "palmerpenguins")),
  comment = "#>"
)
```

```{r attach}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
```

## Use of duckplyr for individual data frames

To enable duckplyr **for individual data frames instead of session wide**, 

- do **not** load duckplyr with `library()`.
- use `duckplyr::as_duckplyr_tibble()` as the first step in your pipe, without attaching the package.

This example illustrates usage of duckplyr for individual data frames.

Use `duckplyr::as_duckplyr_tibble()` to enable processing with duckdb:

```{r}
out <-
  palmerpenguins::penguins %>%
  # CAVEAT: factor columns are not supported yet
  mutate(across(where(is.factor), as.character)) %>%
  duckplyr::as_duckplyr_tibble() %>%
  mutate(bill_area = bill_length_mm * bill_depth_mm) %>%
  summarize(.by = c(species, sex), mean_bill_area = mean(bill_area)) %>%
  filter(species != "Gentoo")
```

The result is a tibble, with its own class.

```{r}
class(out)
names(out)
```

duckdb is responsible for eventually carrying out the operations.
Despite the late filter, the summary is not computed for the Gentoo species.

```{r}
out %>%
  explain()
```

All data frame operations are supported.
Computation happens upon the first request.

```{r}
out$mean_bill_area
```

After the computation has been carried out, the results are available immediately:

```{r}
out
```

## Extensibility

duckplyr also defines a set of generics that provide a low-level implementer's interface for dplyr's high-level user interface.
Other packages may then implement methods for those generics.

```{r extensibility}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
library(duckplyr)
```


```{r overwrite, echo = FALSE}
methods_overwrite()
```

```{r extensibility2}
# Create a relational to be used by examples below
new_dfrel <- function(x) {
  stopifnot(is.data.frame(x))
  new_relational(list(x), class = "dfrel")
}
mtcars_rel <- new_dfrel(mtcars[1:5, 1:4])

# Example 1: return a data.frame
rel_to_df.dfrel <- function(rel, ...) {
  unclass(rel)[[1]]
}
rel_to_df(mtcars_rel)

# Example 2: A (random) filter
rel_filter.dfrel <- function(rel, exprs, ...) {
  df <- unclass(rel)[[1]]

  # A real implementation would evaluate the predicates defined
  # by the exprs argument
  new_dfrel(df[sample.int(nrow(df), 3, replace = TRUE), ])
}

rel_filter(
  mtcars_rel,
  list(
    relexpr_function(
      "gt",
      list(relexpr_reference("cyl"), relexpr_constant("6"))
    )
  )
)

# Example 3: A custom projection
rel_project.dfrel <- function(rel, exprs, ...) {
  df <- unclass(rel)[[1]]

  # A real implementation would evaluate the expressions defined
  # by the exprs argument
  new_dfrel(df[seq_len(min(3, base::ncol(df)))])
}

rel_project(
  mtcars_rel,
  list(relexpr_reference("cyl"), relexpr_reference("disp"))
)

# Example 4: A custom ordering (eg, ascending by mpg)
rel_order.dfrel <- function(rel, exprs, ...) {
  df <- unclass(rel)[[1]]

  # A real implementation would evaluate the expressions defined
  # by the exprs argument
  new_dfrel(df[order(df[[1]]), ])
}

rel_order(
  mtcars_rel,
  list(relexpr_reference("mpg"))
)

# Example 5: A custom join
rel_join.dfrel <- function(left, right, conds, join, ...) {
  left_df <- unclass(left)[[1]]
  right_df <- unclass(right)[[1]]

  # A real implementation would evaluate the expressions
  # defined by the conds argument,
  # use different join types based on the join argument,
  # and implement the join itself instead of relaying to left_join().
  new_dfrel(dplyr::left_join(left_df, right_df))
}

rel_join(new_dfrel(data.frame(mpg = 21)), mtcars_rel)

# Example 6: Limit the maximum rows returned
rel_limit.dfrel <- function(rel, n, ...) {
  df <- unclass(rel)[[1]]

  new_dfrel(df[seq_len(n), ])
}

rel_limit(mtcars_rel, 3)

# Example 7: Suppress duplicate rows
#  (ignoring row names)
rel_distinct.dfrel <- function(rel, ...) {
  df <- unclass(rel)[[1]]

  new_dfrel(df[!duplicated(df), ])
}

rel_distinct(new_dfrel(mtcars[1:3, 1:4]))

# Example 8: Return column names
rel_names.dfrel <- function(rel, ...) {
  df <- unclass(rel)[[1]]

  names(df)
}

rel_names(mtcars_rel)
```


