---
title: "Selective use of duckplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{30 Selective use of duckplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)
  x <- gsub("─", "-", x)
  x
}

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || (getRversion() >= "4.1" && rlang::is_installed(c("conflicted", "nycflights13"))),
  comment = "#>"
)

Sys.setenv(DUCKPLYR_FALLBACK_COLLECT = 0)
```

```{r attach}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
```

This vignette demonstrates how to use duckplyr selectively, for individual data frames or for other packages.

## External data frame

To enable duckplyr for individual data frames instead of session-wide,

- do *not* load duckplyr with `library()`.
- use `duckplyr::as_duckdb_tibble()` as the first step in your pipe, without attaching the package.

```{r}
lazy <-
  duckplyr::flights_df() |>
  duckplyr::as_duckdb_tibble() |>
  filter(!is.na(arr_delay), !is.na(dep_delay)) |>
  mutate(inflight_delay = arr_delay - dep_delay) |>
  summarize(
    .by = c(year, month),
    mean_inflight_delay = mean(inflight_delay),
    median_inflight_delay = median(inflight_delay),
  ) |>
  filter(month <= 6)
```

The result is a tibble, with its own class.

```{r}
class(lazy)
names(lazy)
```

DuckDB is responsible for eventually carrying out the operations.
Despite the filter coming very late in the pipeline, it is applied to the raw data.

```{r}
lazy |>
  explain()
```

All data frame operations are supported.
Computation happens upon the first request.

```{r}
lazy$mean_inflight_delay
```

After the computation has been carried out, the results are preserved and available immediately:

```{r}
lazy
```

## Own data

Construct duckplyr frames directly with `duckplyr::duckdb_tibble()`:

```{r}
data <- duckplyr::duckdb_tibble(
  x = 1:10,
  y = 5,
  z = letters[1:10]
)
data
```


## In other packages

Like other dependencies, duckplyr must be declared in the `DESCRIPTION` file and optionally imported in the `NAMESPACE` file.
Because duckplyr does not import dplyr, it is necessary to import both packages.
The recipe below shows how to achieve this with the usethis package.

- Add dplyr as a dependency with `usethis::use_package("dplyr")`
- Add duckplyr as a dependency with `usethis::use_package("duckplyr")`
- In your code, use a pattern like `data |> duckplyr::as_duckdb_tibble() |> dplyr::filter(...)`
- To avoid the package prefix:
    - Import the duckplyr function with `usethis::use_import_from("duckplyr", "as_duckdb_tibble")`
    - Import the dplyr function with `usethis::use_import_from("dplyr", "filter")`

Learn more about usethis at <https://usethis.r-lib.org/>.


## Funneling

The default mode for `as_duckdb_tibble()` and `duckdb_tibble()` is unfunneled.
This means that the data frame operations are carried out by DuckDB when possible, and also available as data frames upon first request.
Use `as_duckdb_tibble(funnel = TRUE)` or `duckdb_tibble(.funnel = TRUE)` to avoid materializing intermediate data and to ensure that all operations are carried out by DuckDB or fail.
Funneling can also limit the number of rows or cells that are materialized.

```{r}
data <- duckplyr::duckdb_tibble(x = 1:5, .funnel = c(rows = 3))
data
```

```{r error = TRUE}
nrow(data)
```

Learn more about funneling in `vignette("funnel")`, and about the translation employed by duckplyr in `vignette("limits")`.
