---
title: "Use of duckplyr in other packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{developers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)
  x <- gsub("─", "-", x)
  x
}

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || (getRversion() >= "4.1" && rlang::is_installed(c("conflicted", "nycflights13"))),
  comment = "#>"
)

Sys.setenv(DUCKPLYR_FALLBACK_COLLECT = 0)
```

```{r attach}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
```

## Use of duckplyr for individual data frames

To enable duckplyr **for individual data frames instead of session wide**, 

- do **not** load duckplyr with `library()`.
- use `duckplyr::as_duck_tbl()` as the first step in your pipe, without attaching the package.

```{r}
eager <-
  duckplyr::flights_df() |>
  duckplyr::as_duck_tbl() |>
  filter(!is.na(arr_delay), !is.na(dep_delay)) |>
  mutate(inflight_delay = arr_delay - dep_delay) |>
  summarize(
    .by = c(year, month), 
    mean_inflight_delay = mean(inflight_delay),
    median_inflight_delay = median(inflight_delay),
  ) |>
  filter(month <= 6)
```

The result is a tibble, with its own class.

```{r}
class(eager)
names(eager)
```

DuckDB is responsible for eventually carrying out the operations.
Despite the late filter, the summary is not computed for the months in the second half of the year.

```{r}
eager |>
  explain()
```

All data frame operations are supported.
Computation happens upon the first request.

```{r}
eager$mean_inflight_delay
```

After the computation has been carried out, the results are preserved and available immediately:

```{r}
eager
```

## Eager and lazy modes

The default mode for `as_duck_tbl()` is eager.
This allows applying all data frame operations on the results, including column subsetting or retrieving the number of rows.
In addition, if an operation cannot be carried out by duckdb, the dplyr fallback is used transparently.
Use `.lazy = TRUE` to ensure that all operations are carried out by DuckDB, or fail.
This is also the default for the ingestion functions such as `duck_parquet()`.

```{r}
lazy <-
  duckplyr::flights_df() |>
  duckplyr::as_duck_tbl(.lazy = TRUE)
```

Columns or the row count cannot be accessed directly in this mode:

```{r error = TRUE}
nrow(lazy)
```

Also, operations that are not (yet) supported will fail:

```{r error = TRUE}
lazy |>
  mutate(inflight_delay = arr_delay - dep_delay) |>
  summarize(
    .by = c(year, month), 
    mean_inflight_delay = mean(inflight_delay, na.rm = TRUE),
    median_inflight_delay = median(inflight_delay, na.rm = TRUE),
  )
```

See `vignette("limits")` for current limitations, and the contributing guide for how to add support for additional operations.

## Extensibility

duckplyr also defines a set of generics that provide a low-level implementer's interface for dplyr's high-level user interface.
Other packages may then implement methods for those generics.

```{r extensibility}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
library(duckplyr)
```


```{r overwrite, echo = FALSE}
methods_overwrite()
```

```{r extensibility2}
# Create a relational to be used by examples below
new_dfrel <- function(x) {
  stopifnot(is.data.frame(x))
  new_relational(list(x), class = "dfrel")
}
mtcars_rel <- new_dfrel(mtcars[1:5, 1:4])

# Example 1: return a data.frame
rel_to_df.dfrel <- function(rel, ...) {
  unclass(rel)[[1]]
}
rel_to_df(mtcars_rel)

# Example 2: A (random) filter
rel_filter.dfrel <- function(rel, exprs, ...) {
  df <- unclass(rel)[[1]]

  # A real implementation would evaluate the predicates defined
  # by the exprs argument
  new_dfrel(df[sample.int(nrow(df), 3, replace = TRUE), ])
}

rel_filter(
  mtcars_rel,
  list(
    relexpr_function(
      "gt",
      list(relexpr_reference("cyl"), relexpr_constant("6"))
    )
  )
)

# Example 3: A custom projection
rel_project.dfrel <- function(rel, exprs, ...) {
  df <- unclass(rel)[[1]]

  # A real implementation would evaluate the expressions defined
  # by the exprs argument
  new_dfrel(df[seq_len(min(3, base::ncol(df)))])
}

rel_project(
  mtcars_rel,
  list(relexpr_reference("cyl"), relexpr_reference("disp"))
)

# Example 4: A custom ordering (eg, ascending by mpg)
rel_order.dfrel <- function(rel, exprs, ...) {
  df <- unclass(rel)[[1]]

  # A real implementation would evaluate the expressions defined
  # by the exprs argument
  new_dfrel(df[order(df[[1]]), ])
}

rel_order(
  mtcars_rel,
  list(relexpr_reference("mpg"))
)

# Example 5: A custom join
rel_join.dfrel <- function(left, right, conds, join, ...) {
  left_df <- unclass(left)[[1]]
  right_df <- unclass(right)[[1]]

  # A real implementation would evaluate the expressions
  # defined by the conds argument,
  # use different join types based on the join argument,
  # and implement the join itself instead of relaying to left_join().
  new_dfrel(dplyr::left_join(left_df, right_df))
}

rel_join(new_dfrel(data.frame(mpg = 21)), mtcars_rel)

# Example 6: Limit the maximum rows returned
rel_limit.dfrel <- function(rel, n, ...) {
  df <- unclass(rel)[[1]]

  new_dfrel(df[seq_len(n), ])
}

rel_limit(mtcars_rel, 3)

# Example 7: Suppress duplicate rows
#  (ignoring row names)
rel_distinct.dfrel <- function(rel, ...) {
  df <- unclass(rel)[[1]]

  new_dfrel(df[!duplicated(df), ])
}

rel_distinct(new_dfrel(mtcars[1:3, 1:4]))

# Example 8: Return column names
rel_names.dfrel <- function(rel, ...) {
  df <- unclass(rel)[[1]]

  names(df)
}

rel_names(mtcars_rel)
```


