---
title: "duckplyr"
output: rmarkdown::html_vignette
author: Maëlle Salmon
vignette: >
  %\VignetteIndexEntry{00 Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(conflicts.policy = list(warn = FALSE))
```

```{r setup}
library(duckplyr)
```

## Design principles

The duckplyr package uses **DuckDB under the hood** but is also a **drop-in replacement for dplyr**.
These two facts create a tension:

-   When using dplyr, we are not used to explicitly collect results: the data.frames are eager by default.
    Adding a `collect()` step by default would confuse users and make "drop-in replacement" an exaggeration.
    Therefore, _duckplyr needs eagerness_!

-   The whole advantage of using DuckDB under the hood is letting DuckDB optimize computations, like dtplyr does with data.table.
    _Therefore, duckplyr needs laziness_!

As a consequence, duckplyr is lazy on the inside for all DuckDB operations but eager on the outside, thanks to [ALTREP](https://duckdb.org/2024/04/02/duckplyr.html#eager-vs-lazy-materialization), a powerful R feature that among other things supports **deferred evaluation**.

> "ALTREP allows R objects to have different in-memory representations, and for custom code to be executed whenever those objects are accessed."

If the duckplyr data.frame is accessed by...

-   duckplyr, then the operations continue to be lazy (until a call to `collect.duckplyr_df()` for instance).
-   not duckplyr (say, ggplot2), then a special callback is executed, allowing materialization of the data frame.

Therefore, duckplyr can be **both lazy** (within itself) **and not lazy** (for the outside world).

Now, the default materialization can be problematic if dealing with large data: what if the materialization eats up all memory?
Therefore, the duckplyr package has a **safeguard called prudence** with three levels.

- `"lavish"`: automatically materialize _regardless of size_,
- `"frugal"`: _never_ automatically materialize,
- `"thrifty"`: automatically materialize _up to a maximum size of 1 million cells_.

By default, duckplyr frames are _lavish_, but duckplyr frames created from Parquet data (presumedly large) are _thrifty_.

## How to use duckplyr

### For normal sized data (instead of dplyr)

To replace dplyr with duckplyr, you can either

- load duckplyr and then keep your pipeline as is.

```r
library(conflicted)
library(duckplyr)
conflict_prefer("filter", "dplyr", quiet = TRUE)
```

- convert individual data.frames to duck frames which allows you to control their automatic materialization parameters. To do that, you use conversion functions like `duckdb_tibble()` or `as_duckdb_tibble()`, or ingestion functions like `read_csv_duckdb()`.

In both cases, if an operation cannot be performed by duckplyr (see `vignettes("limits")`), it will be outsourced to dplyr.
You can choose to be informed about fallbacks to dplyr, see `?fallback_config`.
You can disable fallbacks by turning off automatic materialization.
In that case, if an operation cannot be performed by duckplyr, your code will error.

### For large data (instead of dbplyr)

With large datasets, you want:

- input data in an efficient format, like Parquet files. Therefore you might input data using `read_parquet_duckdb()`.
- efficient computation, which duckplyr provides via DuckDB's holistic optimization, without your having to use another syntax than dplyr.
- the output to not clutter all the memory. Therefore you can make use of these features:
    - prudence (see `vignette("funnel")`) to disable automatic materialization completely or to disable automatic materialization up to a certain output size.
    - computation to files using  `compute_parquet()` or `compute_csv()`.



A drawback of analyzing large data with duckplyr is that the limitations of duckplyr (unsupported verbs or data types, see `vignette("limits")`) won't be compensated by fallbacks since fallbacks to dplyr necessitate putting data into memory.

## How to improve duckplyr

- telemetry
- report issues, contribute
