---
title: "Fallback to dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Fallback}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)
  x <- gsub("─", "-", x)
  x
}

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || (getRversion() >= "4.1" && rlang::is_installed(c("conflicted", "dbplyr", "nycflights13"))),
  comment = "#>"
)

options(conflicts.policy = list(warn = FALSE))

Sys.setenv(DUCKPLYR_FALLBACK_COLLECT = 0)
```

This article details the fallback mechanism in duckplyr, which allows support for all dplyr verbs and R functions.


```{r attach}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
```


## Introduction

The duckplyr package aims at providing a fully compatible drop-in replacement for dplyr.
To achieve this, only a carefully selected subset of dplyr's operations, R functions, and R data types are implemented.
Whenever a request cannot be handled by DuckDB, duckplyr falls back to dplyr.

## DuckDB operation

The following operation is supported by duckplyr:

```{r}
duckdb <-
  duckplyr::duckdb_tibble(a = 1:3) %>%
  arrange(desc(a)) %>%
  mutate(b = a + 1) %>%
  select(-a)
```

The `explain()` function shows what happens under the hood:

```{r}
duckdb %>%
  explain()
```

The plan shows three operations: a data frame scan (the input), a sort operation, and a projection (adding the `b` column and removing the `a` column).
Because each operation is supported by DuckDB, the resulting object contains a plan for the entire pipeline.
The plan is only executed when the data is needed.

## Relation objects

DuckDB accepts a tree of interconnected _relation objects_ as input.
Each relation object represents a logical step of the execution plan.
The duckplyr package translates dplyr verbs into relation objects.

The `last_rel()` function shows the last relation that has been materialized:

```{r}
duckplyr::last_rel()
```

It is NULL because nothing has been computed yet.
Converting the object to a data frame triggers the computation:

```{r}
duckdb %>% collect()
duckplyr::last_rel()
```

The `last_rel()` function now shows a relation that describes logical plan for executing the whole pipeline.

## Functionality not supported by DuckDB

Using a custom function with a side effect is not supported by DuckDB and triggers a dplyr fallback:

```{r}
verbose_plus_one <- function(x) {
  message("Adding one to ", paste(x, collapse = ", "))
  x + 1
}

fallback <-
  duckplyr::duckdb_tibble(a = 1:3) %>%
  arrange(desc(a)) %>%
  mutate(b = verbose_plus_one(a)) %>%
  select(-a)
```

The `verbose_plus_one()` function is not supported by DuckDB, so the `mutate()` step is handled by dplyr and already executed when the pipeline is defined.
This is confirmed by the `last_rel()` function:

```{r}
duckplyr::last_rel()
```

Only the `arrange()` step is executed by DuckDB.
Because the dplyr implementation of `mutate()` needs the data before it can proceed, the data is first converted to a data frame, and this triggers the materialization of the first step.

The `explain()` function also confirms indirectly that at least a part of the operation is handled by dplyr:

```{r}
fallback %>%
  explain()
```

The final plan now consists of two operations: a data frame scan, and a projection (removing the `a` column).
The data frame scan is the result of the `mutate()` step, which at this stage already has been executed by dplyr.

Converting the final object to a data frame triggers the rest of the computation:

```{r}
fallback %>% collect()
duckplyr::last_rel()
```

The `last_rel()` function confirms that only the final `select()` is handled by DuckDB again.

## Conclusion

The fallback mechanism in duckplyr allows for a seamless integration of dplyr verbs and R functions that are not supported by DuckDB.
It is transparent to the user and only triggers when necessary.
With small or medium-sized data sets, it will not even be noticeable in most settings.

See `vignette("large")` for techniques for working with large data, `vignette("limits")` for the currently implementated translations, `vignette("prudence")` for details on controlling fallback behavior, and `vignette("telemetry")` for the automatic reporting of fallback situations.
