---
title: "Memory protection: Funneling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10 Memory protection: Funneling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)
  x <- gsub("─", "-", x)
  x
}

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || (getRversion() >= "4.1" && rlang::is_installed(c("conflicted", "nycflights13"))),
  comment = "#>"
)

Sys.setenv(DUCKPLYR_FALLBACK_COLLECT = 0)
```

This vignette discusses eager and lazy computation, and funneling.

```{r attach}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
```

## Introduction

Data frames backed by duckplyr, with class `"duckplyr_df"`, behave as regular data frames in almost all respects.
In particular, direct column access like `df$x`, or retrieving the number of rows with `nrow()`, works identically.
Conceptually, duckplyr frames are "eager": from a user's perspective, they behave like regular data frames.

```{r}
df <-
  duckplyr::duckdb_tibble(x = 1:5) |>
  mutate(y = x + 1)
df
class(df)
df$y
nrow(df)
```

Under the hood, two key differences provide improved performance and usability:
lazy materialization and funneling.


## Eager and lazy computation

For a duckplyr frame that is the result of a dplyr operation, accessing column data or retrieving the number of rows will trigger a computation that is carried out by DuckDB, not dplyr.
In this sense, duckplyr frames are also "lazy": the computation is deferred until the last possible moment, allowing DuckDB to optimize the whole pipeline.

### Example

This is explained in the following example that computes the mean arrival delay for flights departing from Newark airport (EWR) by day and month:

```{r}
flights <- duckplyr::flights_df()

flights_duckdb <-
  flights |>
  duckplyr::as_duckdb_tibble()

system.time(
  mean_arr_delay_ewr <-
    flights_duckdb |>
    filter(origin == "EWR", !is.na(arr_delay)) |>
    summarize(
      .by = month,
      mean_arr_delay = mean(arr_delay),
      min_arr_delay = min(arr_delay),
      max_arr_delay = max(arr_delay),
      median_arr_delay = median(arr_delay),
    )
)
```

Setting up the pipeline is fast, the size of the data does not affect the setup costs.
Because the computation is deferred, DuckDB can optimize the whole pipeline, which can be seen in the output below:

```{r}
mean_arr_delay_ewr |>
  explain()
```

The first step is to prune the unneeded columns, only `origin`, `month`, and `arr_delay` are kept.
The result becomes available when accessed:

```{r}
system.time(mean_arr_delay_ewr$mean_arr_delay[[1]])
```

### Comparison

The functionality is similar to lazy tables in dbplyr and lazy frames in dtplyr. However, the behavior is different: at the time of writing, the internal structure of a lazy table or frame is different from a data frame, and columns cannot be accessed directly.

|             | **Eager** 😃 | **Lazy** 😴 |
|-------------|:------------:|:-----------:|
| **dplyr**   |      ✅      |             |
| **dbplyr**  |              |      ✅     |
| **dtplyr**  |              |      ✅     |
| **duckplyr**|      ✅      |      ✅     |

In contrast, with dplyr, each intermediate step and also the final result is a proper data frame, and computed right away, forfeiting the opportunity for optimization:

```{r}
system.time(
  flights |>
    filter(origin == "EWR", !is.na(arr_delay)) |>
    summarize(
      .by = c(month, day),
      mean_arr_delay = mean(arr_delay),
      min_arr_delay = min(arr_delay),
      max_arr_delay = max(arr_delay),
      median_arr_delay = median(arr_delay),
    )
)
```


## Funneling

Being both "eager" and "lazy" at the same time introduces a challenge:
it is too easy to accidentally trigger computation,
which may be prohibitive if an intermediate result is too large.
This is where funneling comes in.


### Concept

For unfunneled duckplyr frames, as in the two previous examples the underlying DuckDB computation is carried out upon the first request.
Once the results are computed, they are cached and subsequent requests are fast.
This is a good choice for small to medium-sized data, where DuckDB can provide a nice speedup but materializing the data is affordable at any stage.
This is the default for `duckdb_tibble()` and `as_duckdb_tibble()`.

For funneled duckplyr frames, accessing a column or requesting the number of rows triggers an error.
This is a good choice for large data sets where the cost of materializing the data may be prohibitive due to size or computation time, and the user wants to control when the computation is carried out.


### Example

The example below demonstrates the use of funneled duckplyr frames.

```{r}
flights_funneled <-
  flights |>
  duckplyr::as_duckdb_tibble(funnel = "frugal")
```

In this example, `flights_funneled` is a funneled duckplyr frame.
The data can be displayed, and column names and types can be accessed.

```{r}
flights_funneled
names(flights_funneled)[1:10]
class(flights_funneled)
class(flights_funneled[[1]])
```

On the other hand, accessing a column or requesting the number of rows triggers an error:

```{r error = TRUE}
nrow(flights_funneled)
flights_funneled[[1]]
```


### Enforcing DuckDB operation

For operations not supported by duckplyr, the original dplyr implementation is used as a fallback.
As the original dplyr implementation accesses columns directly, the data must be materialized before a fallback can be executed.
Therefore, funneled frames allow you to check that all operations are supported by DuckDB: for a funneled frame, fallbacks to dplyr are not possible.

```{r error = TRUE}
flights_funneled |>
  group_by(origin) |>
  summarize(n = n()) |>
  ungroup()
```

The same pipeline with an unfunneled frame works, but the computation is carried out by dplyr:

```{r}
flights_funneled |>
  duckplyr::as_duckdb_tibble(funnel = "lavish") |>
  group_by(origin) |>
  summarize(n = n()) |>
  ungroup()
```

By using operations supported by duckplyr and avoiding fallbacks as much as possible, your pipelines will be executed by DuckDB in an optimized way.
See `?fallback` for details on fallbacks, and `vignette("limits")` for the operations supported by duckplyr.


### From funneled to unfunneled

A funneled duckplyr frame can be converted to an unfunneled one with `as_duckdb_tibble(funnel = "lavish")`.
The `collect.duckplyr_df()` method triggers computation and converts to a plain tibble.
The difference between the two is the class of the returned object:

```{r}
flights_funneled |>
  duckplyr::as_duckdb_tibble(funnel = "lavish") |>
  class()

flights_funneled |>
  collect() |>
  class()
```

The same behavior is achieved with `as_tibble()` and `as.data.frame()`:

```{r}
flights_funneled |>
  as_tibble() |>
  class()

flights_funneled |>
  as.data.frame() |>
  class()
```

See `vignette("large")` for techniques for working with large data sets.

### Comparison

Funneled duckplyr frames behave like lazy tables in dbplyr and lazy frames in dtplyr: the computation only starts when you *explicitly* request it with `collect.duckplyr_df()` or through other means.
However, funneled duckplyr frames can be converted to unfunneled ones at any time, and vice versa.
In dtplyr and dbplyr, there are no unfunneled frames: collection always needs to be explicit.


## Partial funneling

Partial funneling is a compromise between funneled and unfunneled.
Materialization is allowed for data up to a certain size, measured in cells (values) and rows in the resulting data frame.

```{r}
nrow(flights)
flights_partial <-
  flights |>
  duckplyr::as_duckdb_tibble(funnel = "thrifty")
```

With this setting, the data is materialized only if the result has fewer than 1,000,000 cells (rows multiplied by columns).

```{r error = TRUE}
flights_partial |>
  select(origin) |>
  nrow()
```

The original input is too large to be materialized, so the operation fails.
On the other hand, the result after aggregation is small enough to be materialized:

```{r}
flights_partial |>
  count(origin) |>
  nrow()
```

Partial funneling is a good choice for data sets where the cost of materializing the data is prohibitive only for large results.
This is the default for the ingestion functions like `read_parquet_duckdb()`.
See `vignette("large")` for more details on working with large data sets.
