---
title: "Large data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01 Large data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)
  x <- gsub("─", "-", x)
  x
}

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || (getRversion() >= "4.1" && rlang::is_installed(c("conflicted", "dbplyr", "nycflights13"))),
  comment = "#>"
)

options(conflicts.policy = list(warn = FALSE))

Sys.setenv(DUCKPLYR_FALLBACK_COLLECT = 0)
```

Working with large datasets in R can be challenging, especially when performance and memory constraints are a concern.
The duckplyr package, built on top of DuckDB, offers a powerful solution by enabling efficient data manipulation using familiar dplyr syntax.
This article explores strategies for handling large datasets with duckplyr, covering ingestion, materialization of intermediate and final results, and good practice.

```{r attach}
library(conflicted)
library(duckplyr)
conflict_prefer("filter", "dplyr")
```


## Working with large data is challenging...

But duckplyr can help.

When working with large data, that could be larger than the available memory, we need

- **Input data to be ingested from specific formats such as Parquet, and to not waste all the memory.** Ingestion functions in duckplyr can read data from CSV, JSON, Parquet, into a frame that by default cannot be materialized if its size is larger than a threshold.

- **Efficient computation.** Speed is the whole point of duckplyr, that gives you the power of DuckDB without even having to learn its syntax.

- **Results to also not waste memory.** The duckplyr package lets you control whether automatic materialization will happen, and lets you compute results to a file rather than memory.

## Inputting data to duckplyr


### Conversion functions

The `duckdb_tibble()` function creates a duckplyr data frame from vectors:

```{r}
df <- duckdb_tibble(x = 1:3, y = letters[1:3])
df
```

The `duckdb_tibble()` function is a drop-in replacement for `tibble()`, and can be used in the same way.

Similarly, `as_duckdb_tibble()` can be used to convert a data frame or another object to a duckplyr data frame:

```{r}
flights_df() |>
  as_duckdb_tibble()
```

Existing code that uses DuckDB via dbplyr can also take advantage.
The following code creates a DuckDB connection and writes a data frame to a table:

```{r}
path_duckdb <- tempfile(fileext = ".duckdb")
con <- DBI::dbConnect(duckdb::duckdb(path_duckdb))
DBI::dbWriteTable(con, "data", data.frame(x = 1:3, y = letters[1:3]))

dbplyr_data <- tbl(con, "data")
dbplyr_data

dbplyr_data |>
  explain()
```

The `explain()` output shows that the data is actually coming from a DuckDB table.
The `as_duckdb_tibble()` function can then be used to seamlessly convert the data to a duckplyr frame:

```{r}
dbplyr_data |>
  as_duckdb_tibble()

dbplyr_data |>
  as_duckdb_tibble() |>
  explain()
```

This only works for DuckDB connections.
For other databases, turn the data into an R data frame or export it to a file before using `as_duckdb_tibble()`.

```{r}
DBI::dbDisconnect(con)
```

For other common cases, the `duckdb_tibble()` function fails with a helpful error message:

```{r error = TRUE}
duckdb_tibble(a = 1) |>
  group_by(a) |>
  as_duckdb_tibble()
duckdb_tibble(a = 1) |>
  rowwise() |>
  as_duckdb_tibble()
readr::read_csv("a\n1", show_col_types = FALSE) |>
  as_duckdb_tibble()
```

In all cases, `as_tibble()` can be used to proceed with the existing code.

### Ingestion functions

DuckDB supports data ingestion from CSV, Parquet, and JSON files.
The `read_csv_duckdb()` function accepts a file path and returns a duckplyr frame.

```{r}
path_csv_1 <- tempfile(fileext = ".csv")
writeLines("x,y\n1,a\n2,b\n3,c", path_csv_1)
read_csv_duckdb(path_csv_1)
```

Reading multiple files is also supported:

```{r}
path_csv_2 <- tempfile(fileext = ".csv")
writeLines("x,y\n4,d\n5,e\n6,f", path_csv_2)
read_csv_duckdb(c(path_csv_1, path_csv_2))
```

The `options` argument can be used to control the reading.

Similarly, the `read_parquet_duckdb()` and `read_json_duckdb()` functions can be used to read Parquet and JSON files, respectively.

For reading from HTTPS or S3 URLs, the [httpfs extension](https://duckdb.org/docs/extensions/httpfs/overview.html) must be installed and loaded in each session.

```{r}
db_exec("INSTALL httpfs")
db_exec("LOAD httpfs")
```

Installation is fast if the extension is already installed.
Once loaded, the `read_csv_duckdb()`, `read_parquet_duckdb()`, and `read_json_duckdb()` functions can be used with URLs:

```{r}
url <- "https://blobs.duckdb.org/flight-data-partitioned/Year=2024/data_0.parquet"
flights_parquet <- read_parquet_duckdb(url)
flights_parquet
```

In all cases, the data is read lazily: only the metadata is read initially, and the data is read as required.
This means that data can be read from files that are larger than the available RAM.
The Parquet format is particularly efficient for this purpose, as it stores data in a columnar format and allows reading only the columns that are required.
See `vignette("prudence")` for more details on the concept of lazy data.

### From DuckDB

In addition to `as_duckdb_tibble()`, arbitrary DuckDB queries can be executed and the result can be converted to a duckplyr frame.
For this, [attach](https://duckdb.org/docs/sql/statements/attach.html) an existing DuckDB database first:

```{r}
sql_attach <- paste0(
  "ATTACH DATABASE '",
  path_duckdb,
  "' AS external (READ_ONLY)"
)
db_exec(sql_attach)
```

Then, use `read_sql_duckdb()` to execute a query and return a duckplyr frame:

```{r}
read_sql_duckdb("SELECT * FROM external.data")
```

## Materialization

In dbplyr, `compute()` is used to materialize a lazy table in a temporary table on the database, and `collect()` is used to bring the data into R memory.
This interface works exactly the same in duckplyr:

```{r}
simple_data <-
  duckdb_tibble(a = 1) |>
  mutate(b = 2)

simple_data |>
  explain()

simple_data_computed <-
  simple_data |>
  compute()
```

The `compute.duckplyr_df()` function returns a duckplyr frame that is materialized in a temporary table.
The return value of the function is a duckplyr frame that can be used in further computations.
The materialization is done in a temporary table, so the data is not persisted after the session ends:

```{r}
simple_data_computed |>
  explain()
```

The `collect()` function brings the data into R memory and returns a plain tibble:

```{r}
duckdb_tibble(a = 1) |>
  mutate(b = 2) |>
  collect()
```

### To files

To materialize data in a persistent file, the `compute_csv()` and `compute_parquet()` functions can be used.
The `compute_csv()` function writes the data to a CSV file:

```{r}
path_csv_out <- tempfile(fileext = ".csv")
duckdb_tibble(a = 1) |>
  mutate(b = 2) |>
  compute_csv(path_csv_out)

writeLines(readLines(path_csv_out))
```

The `compute_parquet()` function writes the data to a Parquet file:

```{r}
path_parquet_out <- tempfile(fileext = ".parquet")
duckdb_tibble(a = 1) |>
  mutate(b = 2) |>
  compute_parquet(path_parquet_out) |>
  explain()
```

Just like with `compute.duckplyr_df()`, the return value of `compute_csv()` and `compute_parquet()` is a duckplyr frame that uses the created CSV or Parquet file and can be used in further computations.
At the time of writing, direct JSON export is not supported.

## Memory protection

### Control automatic materialization

See `vignette("prudence")`.

### Memory usage

Computations carried out by DuckDB allocate RAM in the context of the R process.
This memory separate from the memory used by R objects, and is managed by DuckDB.
Limit the memory used by DuckDB by setting a pragma:

```{r}
read_sql_duckdb("SELECT current_setting('memory_limit') AS memlimit")

db_exec("PRAGMA memory_limit = '1GB'")

read_sql_duckdb("SELECT current_setting('memory_limit') AS memlimit")
```

See [the DuckDB documentation](https://duckdb.org/docs/configuration/overview.html) for other configuration options.

## The big picture

The functions shown in this vignette allow the construction of data transformation pipelines spanning multiple data sources and data that is too large to fit into memory.
Full compatibility with dplyr is provided, so existing code can be used with duckplyr with minimal changes.
The lazy evaluation of duckplyr frames allows for efficient data processing, as only the required data is read from disk.
The materialization functions allow the data to be persisted in temporary tables or files, depending on the use case.

A typical workflow might look like this:

- Prepare all data sources as duckplyr frames: local data frames and files.
- Combine the data sources using dplyr verbs.
- Preview intermediate results as usual: the computation will be faster because only the first few rows are requested.
- To avoid rerunning the whole pipeline all over, use `compute.duckplyr_df()` or `compute_parquet()` to materialize any intermediate result that is too large to fit into memory.
- Collect the final result using `collect.duckplyr_df()` or write it to a file using `compute_csv()` or `compute_parquet()`.

There is a caveat: due to the design of duckplyr, if a dplyr verb is not supported or uses a function that is not supported, the data will be read into memory before being processed further.

By default, if the data pipeline starts with an ingestion function, the data will only be read into memory if it is less than 1 million cells or values in the table:

```{r error = TRUE}
flights_parquet |>
  group_by(Month)
```

Because `group_by()` is not supported, the data will be attempted to read into memory before the `group_by()` operation is executed.
Once the data is small enough to fit into memory, this works transparently.

```{r}
flights_parquet |>
  count(Month, DayofMonth) |>
  group_by(Month)
```

See `vignette("prudence")` for the concepts and mechanisms at play, and `vignette("fallback")` for a detailed explanation of the fallback mechanism.
