---
title: "Translations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{20 Translations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)
  x <- gsub("─", "-", x)
  x
}

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") || (getRversion() >= "4.1" && rlang::is_installed(c("conflicted", "nycflights13", "lubridate"))),
  comment = "#>"
)

Sys.setenv(DUCKPLYR_FALLBACK_COLLECT = 0)
```

This article describes the translations provided by duckplyr for different data types, verbs, and functions within verbs.
If a translation is not provided, duckplyr falls back to dplyr, see `vignette("fallback")` for details.
The translation layer can be bypassed, see `vignette("duckdb")` for details.

```{r attach}
library(conflicted)
library(dplyr)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

## Data types

duckplyr supports the following data types:

- `is.logical()`
- `is.integer()`
- `is.numeric()`
- `is.character()`
- `is.Date()`
- `is.POSIXct()` (with UTC time zone)
- `is.difftime()`

```{r}
duckplyr::duckdb_tibble(
  logical = TRUE,
  integer = 1L,
  numeric = 1.1,
  character = "a",
  Date = as.Date("2025-01-11"),
  POSIXct = as.POSIXct("2025-01-11 19:23:00", tz = "UTC"),
  difftime = as.difftime(1, units = "secs"),
) |>
  compute()
```

Generally, zero-column tibbles are not supported by duckplyr, neither as input nor as a result.

```{r error = TRUE}
duckplyr::duckdb_tibble()
duckplyr::duckdb_tibble(a = 1, .prudence = "stingy") |>
  select(-a)
```

Support for more data types, and passthrough of unknown data types, is planned.
Let's [discuss](https://github.com/tidyverse/duckplyr/discussions/) any additional data types you would like to see supported.


## Verbs

Not all dplyr verbs are implemented within duckplyr.
For unsupported verbs, duckplyr automatically falls back to dplyr.
See `?unsupported` for a list of verbs for which duckplyr does not provide a method.

See the [reference index](https://duckplyr.tidyverse.org/reference/index.html) for a list of verbs with corresponding duckplyr methods.

Let's [discuss](https://github.com/tidyverse/duckplyr/discussions/) any additional verbs you would like to see supported.


## Functions within verbs

For all functions used in dplyr verbs, translations must be provided.
If an expression contains a function for which no translation is provided, duckplyr falls back to dplyr.
With some exceptions, only positional matching is implemented.

As of now, here are the translations provided:

### Parentheses

Implemented: `(`.

Reference: `?Paren`.

```{r}
duckplyr::duckdb_tibble(a = 1, b = 2, c = 3, .prudence = "stingy") |>
  mutate((a + b) * c)
```

### Comparison operators

Implemented: `>`, `>=`, `<`, `<=`, `==`, `!=`.

Reference: `?Comparison`.

```{r}
duckplyr::duckdb_tibble(
  a = c(1, 2, NA),
  b = c(2, NA, 3),
  c = c(NA, 3, 4),
  .prudence = "stingy"
) |>
  mutate(a > b, b != c, c < a, a >= b, b <= c)
```

### Basic arithmetics

Implemented: `+`, `-`, `*`, `/`.

Reference: `?Arithmetic`.

```{r}
duckplyr::duckdb_tibble(a = 1, b = 2, c = 3, .prudence = "stingy") |>
  mutate(a + b, a / b, a - b, a * b)
```

### Math functions

Implemented: `log()`, `log10()`, `abs()`.

Reference: `?Math`.

```{r}
duckplyr::duckdb_tibble(a = 1, b = 2, c = -3, .prudence = "stingy") |>
  mutate(log10(a), log(b), abs(c))
```

### Logical operators

Implemented: `!`, `&`, `|`.

Reference: `?Logic`.

```{r}
duckplyr::duckdb_tibble(a = FALSE, b = TRUE, c = NA, .prudence = "stingy") |>
  mutate(!a, a & b, b | c)
```

### Branching and conversion

Implemented:

- `is.na()`, `as.integer()`
- `dplyr::if_else()`, `dplyr::coalesce()`, `dplyr::case_when()`
- `strftime(x, format)`

```{r}
duckplyr::duckdb_tibble(a = 1, b = NA, .prudence = "stingy") |>
  mutate(is.na(b), if_else(is.na(b), 0, 1), as.integer(b))

duckplyr::duckdb_tibble(a = 23, .prudence = "stingy") |>
  mutate(b = case_when(a == 1 ~ 3, a == 2 ~ 28, .default = 4))

duckplyr::duckdb_tibble(
  a = as.POSIXct("2025-01-11 19:23:46", tz = "UTC"),
  .prudence = "stingy") |>
  mutate(strftime(a, "%H:%M:%S"))
```

### String manipulation

Implemented: `grepl()`, `substr()`, `sub()`, `gsub()`.

```{r}
duckplyr::duckdb_tibble(a = "abbc", .prudence = "stingy") |>
  mutate(grepl("b", a), substr(a, 2L, 3L), sub("b", "B", a), gsub("b", "B", a))
```

### Date manipulation

Implemented: `lubridate::hour()`, `lubridate::minute()`, `lubridate::second()`, `lubridate::wday()`.

```{r}
duckplyr::duckdb_tibble(
  a = as.POSIXct("2025-01-11 19:23:46", tz = "UTC"),
  .prudence = "stingy"
) |>
  mutate(
    hour = lubridate::hour(a),
    minute = lubridate::minute(a),
    second = lubridate::second(a),
    wday = lubridate::wday(a)
  )
```

### Aggregation

Implemented:

- `sum(x, na.rm)`, `dplyr::n()`, `dplyr::n_distinct()`
- `mean(x, na.rm)`, `median(x, na.rm)`, `sd(x, na.rm)`
- `min()`, `max()`, `any()`, `all()`

```{r}
duckplyr::duckdb_tibble(a = 1:3, b = c(1, 2, 2), .prudence = "stingy") |>
  summarize(
    sum(a),
    n(),
    n_distinct(b),
  )

duckplyr::duckdb_tibble(a = 1:3, b = c(1, 2, NA), .prudence = "stingy") |>
  summarize(
    mean(b, na.rm = TRUE),
    median(a),
    sd(b),
  )

duckplyr::duckdb_tibble(a = 1:3, .prudence = "stingy") |>
  summarize(
    min(a),
    max(a),
    any(a > 1),
    all(a > 1),
  )
```

### Shifting

All optional arguments to `dplyr::lag()` and `dplyr::lead()` are supported.

```{r}
duckplyr::duckdb_tibble(a = 1:3, .prudence = "stingy") |>
  mutate(lag(a), lead(a))
duckplyr::duckdb_tibble(a = 1:3, .prudence = "stingy") |>
  mutate(lag(a, 2), lead(a, n = 2))
duckplyr::duckdb_tibble(a = 1:3, .prudence = "stingy") |>
  mutate(lag(a, default = 0), lead(a, default = 4))
duckplyr::duckdb_tibble(a = 1:3, b = c(2, 3, 1), .prudence = "stingy") |>
  mutate(lag(a, order_by = b), lead(a, order_by = b))
```

### Ranking

[Ranking in DuckDB](https://duckdb.org/docs/sql/functions/window_functions.html) is very different from dplyr.
Most functions in DuckDB rank only by the current row number, whereas in dplyr, ranking is done by a column.
It will be difficult to provide translations for the following ranking functions.

- `rank()`, `dplyr::min_rank()`, `dplyr::dense_rank()`
- `dplyr::percent_rank()`, `dplyr::cume_dist()`

Implementing `dplyr::ntile()` is feasible for the `n` argument.
The only ranking function currently implemented is `dplyr::row_number()`.

```{r}
duckplyr::duckdb_tibble(a = c(1, 2, 2, 3), .prudence = "stingy") |>
  mutate(row_number())
```

### Special cases

`$` (`?Extract`) is implemented if the LHS is `.data` or `.env`:

```{r}
b <- 4
duckplyr::duckdb_tibble(a = 1, b = 2, .prudence = "stingy") |>
  mutate(.data$a + .data$b, .env$b)
```

`%in%` (`?match`) is implemented if the RHS is a constant with up to 100 values:

```{r}
duckplyr::duckdb_tibble(a = 1:3, .prudence = "stingy") |>
  mutate(a %in% c(1, 3)) |>
  collect()
duckplyr::last_rel()
```

`dplyr::desc()` is only implemented in the context of `dplyr::arrange()`:

```{r}
duckplyr::duckdb_tibble(a = 1:3, .prudence = "stingy") |>
  arrange(desc(a)) |>
  explain()
```

`suppressWarnings()` is a no-op:

```{r}
duckplyr::duckdb_tibble(a = 1, .prudence = "stingy") |>
  mutate(suppressWarnings(a + 1))
```

### Contributing

Refer to [our contributing guide](https://duckplyr.tidyverse.org/CONTRIBUTING.html#new-translations-for-functions) to learn how to contribute new translations to the package.
Ideally, duckplyr will also support adding custom translations for functions for the duration of the current R session.

## Known incompatibilities

This section tracks known incompatibilities between dplyr and duckplyr.
Changing these is likely to require substantial effort, and might be best addressed by providing new functions with consistent behavior in both dplyr and DuckDB.

### Output order stability

DuckDB does not guarantee order stability for the output.
For performance reasons, duckplyr does not enable output order stability by default.

```{r}
duckplyr::flights_df() |>
  duckplyr::as_duckdb_tibble() |>
  distinct(day) |>
  summarize(paste(day, collapse = " ")) # fallback

duckplyr::flights_df() |>
  distinct(day) |>
  summarize(paste(day, collapse = " "))
```

This can be changed globally with the `DUCKPLYR_OUTPUT_ORDER` environment variable, see `?config` for details.
With this setting, the output order is stable, but the plans are more complicated, and DuckDB needs to do more work.

```{r}
duckplyr::flights_df() |>
  duckplyr::as_duckdb_tibble() |>
  distinct(day) |>
  explain()

withr::with_envvar(
  c(DUCKPLYR_OUTPUT_ORDER = "TRUE"),
  duckplyr::flights_df() |>
    duckplyr::as_duckdb_tibble() |>
    distinct(day) |>
    explain()
)
```

### `sum()`

In duckplyr, this function returns a numeric value also for integers, due to DuckDB's type stability requirement.

```{r}
duckplyr::duckdb_tibble(a = 1:100) |>
  summarize(sum(a))

duckplyr::duckdb_tibble(a = 1:1000000) |>
  summarize(sum(a))

tibble(a = 1:100) |>
  summarize(sum(a))

tibble(a = 1:1000000) |>
  summarize(sum(a))
```

### Empty vectors in aggregate functions

At the time of writing, empty vectors only occur when summarizing an empty table without grouping.
In all cases, duckplyr returns `NA`, and the behavior of dplyr is different:

- `sum()` for an empty vector returns `0`
- `any()` and `all()` return `FALSE`
- `min()` and `max()` return infinity values (with a warning)

```{r}
duckplyr::duckdb_tibble(a = integer(), b = logical()) |>
  summarize(sum(a), any(b), all(b), min(a), max(a))
tibble(a = integer(), b = logical()) |>
  summarize(sum(a), any(b), all(b), min(a), max(a))
```

### `min()` and `max()` for logical input

For completeness, duckplyr returns a logical for `min()` and `max()` when the input is logical, while dplyr returns an integer.

```{r}
duckplyr::duckdb_tibble(a = c(TRUE, FALSE)) |>
  summarize(min(a), max(a))

tibble(a = c(TRUE, FALSE)) |>
  summarize(min(a), max(a))
```

### `n_distinct()` and multiple arguments

This function needs exactly one argument besides the optional `na.rm`. Multiple arguments is not supported.

### `is.na()` and `NaN` values

This function returns `FALSE` for `NaN` values in duckplyr, while it returns `TRUE` in dplyr.

```{r}
duckplyr::duckdb_tibble(a = c(NA, NaN)) |>
  mutate(is.na(a))

tibble(a = c(NA, NaN)) |>
  mutate(is.na(a))
```

### Row names

DuckDB does not support data frames with row names.
When converting a data frame with row names to a duckplyr data frame, the row names are silently stripped.
This is relevant when working with data frames that have row names, such as `mtcars`.

```{r}
# mtcars has character row names
head(rownames(mtcars))
# After conversion, the row names are lost
mtcars |>
  duckplyr::as_duckdb_tibble() |>
  head()
```

To preserve row names, convert them to a column before using duckplyr:

```{r}
mtcars |>
  tibble::rownames_to_column("name") |>
  duckplyr::as_duckdb_tibble() |>
  head()
```

### Other differences

Does the same pipeline give different results with `tibble()` and `duckdb_tibble()`?
We would love to hear about it, please file an [issue](https://github.com/tidyverse/duckplyr/issues/new).
