---
title: "Translations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{translations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Verbs

Not all dplyr verbs are implemented within duckplyr.
For unsupported verbs, duckplyr automatically falls back to dplyr.
See `?not-supported` for a list of verbs for which duckplyr does not provide a method.

See the [reference index](https://duckplyr.tidyverse.org/reference/index.html) for a list of verbs with corresponding duckplyr methods.

Refer to [our contributing guide](https://duckplyr.tidyverse.org/CONTRIBUTING.html#new-translations-for-functions) to learn how to suggest the addition of a new verb.


## Functions within verbs

For all functions used in dplyr verbs, translations must be provided. 

As of now, here are the translations provided:

```{r}
knitr::kable(
  read.csv(system.file("translated.csv", package = "duckplyr"))
)
```

Refer to [our contributing guide](https://duckplyr.tidyverse.org/CONTRIBUTING.html#new-translations-for-functions) to learn how to contribute new translations.
